Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-20 co-regular-EPQ.tex

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection. Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
To prove the $\textit{only if}$ direction, let $e$ be an execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

Given $e\vert_{\preceq p} \sqsubseteq l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. By above lemma we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ into $p$-incomparable values is l. We can see that $u$, $v$ and $w$ are sub-sequence of $u'$, $v'$ an $w'$, respectively. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ does not have restriction to $p$-incomparable values, we can see that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. \qed
\end {proof}

@@

>>>>

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contains only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection. 
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

Thus, let $e$ be an execution such that $e\vert_{\preceq p}$ is linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. By the property above, we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ to values of priority comparable to $p$ is $l$. 
%We can see that $u$, $v$ and $w$ are subsequences of $u'$, $v'$ an $w'$, respectively. 
Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ does not have a condition on values of priority incomparable to $p$, we obtain that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$.
\end {proof}

@@

%{\color {blue}When $\mathsf{MatchedMaxPriority}$ is false for some execution $e$, then there are two possibilities:
%
%\begin{itemize}
%\setlength{\itemsep}{0.5pt}
%\item[-] There exists value $x$, let $e_x$ be obtained from $e$ when all the values of maximal priorities in $e$ but $x$ are removed. Then $\mathsf{MatchedMaxPriority\text{-}Conc}(e_x,x)$ is false.
%
%\item[-] There exists set $X$ of values, let $e'$ be obtained from $e$ when all the values of maximal priorities in $e$ but values in $X$ are removed. Then, although $\mathsf{MatchedMaxPriority\text{-}Conc}(e'_x,x)$ is true for each $x \in X$, there is no value $y$ to make $\mathsf{MatchedMaxPriority\text{-}Conc}(e',y)$ to be true.
%\end{itemize}
%
%It is surprised to see that these two possibilities can be captured by two sub-cases of $\mathsf{MatchedMaxPriority}$, depending on whether is only one value $x$ with maximal priority $p$. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with this condition (i.e., all the values other than $x$ have a priority strictly smaller than $p$) and $\mathsf{MatchedMaxPriority}^{=}$ for the strengthening of the same formula with the negation of this condition (i.e., there exists another value $y$ with priority $p$).
%}
